---
title: "`r if (nzchar(params$scenario)) paste0('Norma Evaluation Summary - ', params$scenario) else 'Norma Evaluation Summary'`"
output: html_document
date: "`r Sys.Date()`"
params:
    datafile: "measurements.csv"
    scenario: ""
---

```{r setup, echo=FALSE, include=FALSE}
library(tidyverse)
library(lubridate)
library(slider)

figure_dimensions <- c(10, 4)

# Load monitoring data
all_data <- read_csv(params$datafile)
start_time <- min(all_data$time, na.rm = TRUE)
if (nrow(all_data) > 0) {
    all_data <- all_data %>%
        mutate(elapsed_sec = as.numeric(time - start_time) / 1e9)
} else {
    # Guard against empty data sets to prevent crashes
    all_data$elapsed_sec <- 0
}

# --- Helper functions ---

# Filter a named metric and convert its value column to numeric.
filter_metric <- function(data, metric_name) {
    data %>%
        dplyr::filter(metric == metric_name) %>%
        mutate(value = as.numeric(value))
}

# ggplot theme with a horizontally centered title.
norma_theme <- function() {
    theme(plot.title = element_text(hjust = 0.5))
}

# Scatter plot of a block metric with a smooth trend line.
# Expects columns: block, value.
block_trend_plot <- function(data, title, ylab) {
    ggplot(data, aes(x = block, y = value)) +
        geom_point() +
        geom_smooth(se = FALSE) +
        ggtitle(title) +
        xlab("Block Height") +
        ylab(ylab) +
        norma_theme()
}

# Line plot of a per-node metric over elapsed time.
# Expects columns: elapsed_sec, value, node.
node_time_plot <- function(data, title, ylab,
                           xlab = "Time",
                           color_label = "Nodes") {
    ggplot(data, aes(
        x = elapsed_sec, y = value,
        group = node, colour = factor(node)
    )) +
        geom_line() +
        ggtitle(title) +
        xlab(xlab) +
        ylab(ylab) +
        labs(colour = color_label) +
        norma_theme()
}

# Plot of a per-node metric over block height.
# Expects columns: block, value, node.
# When smooth = TRUE, draws trend lines instead of scatter points.
node_block_plot <- function(data, title, ylab, smooth = FALSE) {
    p <- ggplot(data, aes(
        x = block, y = value,
        group = node, colour = factor(node)
    )) +
        ggtitle(title) +
        xlab("Block Height") +
        ylab(ylab) +
        labs(colour = "Nodes") +
        norma_theme()

    if (smooth) p + geom_smooth(se = FALSE) else p + geom_point()
}

# Compute a 5-second sliding window transaction rate from accumulated counts.
# data:       rows for a single metric, with numeric `time` and `value` columns
# group_cols: character vector of column names to group by
# Returns a data frame with columns: rate, time, elapsed_sec
compute_sliding_rate <- function(data, group_cols, start_time) {
    result <- data %>%
        mutate(value = as.numeric(value), time = as.numeric(time)) %>%
        group_by(across(all_of(group_cols))) %>%
        arrange(time) %>%
        reframe(
            time = time,
            increment = value - lag(value, default = 0)
        ) %>%
        arrange(time) %>%
        slide_period_dfr(
            as.POSIXct(.$time / 1e9, origin = "1970-01-01"),
            "second",
            .every = 5,
            function(d) {
                summarize(d,
                    rate = sum(increment) / 5,
                    time = (max(time) + min(time)) / 2
                )
            }
        )

    if (nrow(result) > 0) {
        result %>%
            drop_na(rate) %>%
            mutate(elapsed_sec = (time - as.numeric(start_time)) / 1e9)
    } else {
        result %>% mutate(rate = 0, elapsed_sec = 0)
    }
}

# Compute per-observation incremental rate (delta between consecutive values).
# data:       rows for a single metric with elapsed_sec and value columns
# group_cols: character vector of column names to group by
# Returns a data frame with: elapsed_sec, rate, and the grouping columns
compute_incremental_rate <- function(data, group_cols) {
    data %>%
        mutate(value = as.numeric(value)) %>%
        group_by(across(all_of(group_cols))) %>%
        arrange(across(c(all_of(group_cols), "elapsed_sec"))) %>%
        reframe(
            elapsed_sec = elapsed_sec,
            rate = value - lag(value, default = 0)
        )
}
```

This report summarizes the monitoring results of an execution of a Norma scenario.

## Block Metrics

This section covers per-Block metrics which are consistent throughout the network.

### Gas per Block
The following chart shows the Gas spent per block and a trend line showing long-term trends.


```{r gas_spent_per_block, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "BlockGasUsed") %>%
    block_trend_plot("Gas per Block", "Gas spent")
```


### Transactions per Block
The following chart shows the number of Transactions per block and a trend line showing long-term trends.

```{r transactions_per_block, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "BlockNumberOfTransactions") %>%
    block_trend_plot("Transactions per Block", "Number of Transactions")
```


### Block Delay
The following chart shows the time between the earliest reported completion time of block `n-1` and earliest reported completion time block `n` for block `n`.

```{r block_delay_time, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
all_data %>%
    dplyr::filter(metric == "BlockCompletionTime") %>%
    select(block, value) %>%
    mutate(value = as.numeric(value)) %>%
    group_by(block) %>%
    summarise(value = min(value)) %>%
    arrange(block) %>%
    mutate(delta = value - lag(value)) %>%
    drop_na(delta) %>%
    ggplot(aes(x = block, y = delta / 1e9)) +
        geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
        geom_point() +
        geom_smooth(se = FALSE) +
        ggtitle("Block Delay") +
        xlab("Block Height") +
        ylab("Delay [s]") +
        norma_theme()
```


### Gas Rate per Block
The following chart shows the Gas Rate per block and a trend line.

```{r gas_rate_per_block, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "BlockGasRate") %>%
    block_trend_plot("Gas Rate per Block", "Gas per Second")
```


### Gas Base-Fee per Block
The following chart shows the Gas Base-Fee per block and a trend line.

```{r gas_base_fee_per_block, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "BlockGasBaseFee") %>%
    block_trend_plot("Gas Base Fee per Block", "Gas Base Fee")
```

## Network Metrics

This section covers metrics that are network wide properties.

### Network Incoming and Committed Transaction Rates

The following chart shows the total rate of incoming and committed transactions smoothed over 5 second intervals.

```{r throughput_over_time, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
sent_rate <- all_data %>%
    dplyr::filter(metric == "SentTransactions") %>%
    select(time, app, workers, value) %>%
    compute_sliding_rate(c("app", "workers"), start_time)

received_rate <- all_data %>%
    dplyr::filter(metric == "ReceivedTransactions") %>%
    select(time, app, value) %>%
    compute_sliding_rate("app", start_time)

ggplot() +
    geom_line(
        data = sent_rate,
        aes(x = elapsed_sec, y = rate, color = "incomming [5s window]")
    ) +
    geom_line(
        data = received_rate,
        aes(x = elapsed_sec, y = rate, color = "committed [5s window]")
    ) +
    ggtitle("Incomming and Committed Transaction Rates") +
    xlab("Time (s)") +
    ylab("Tx/s") +
    norma_theme()
```


### Number of Nodes in the Network

The number of nodes in the network over time.

```{r nodes_in_network, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "NumberOfNodes") %>%
    ggplot(aes(x = elapsed_sec, y = value)) +
        geom_point() +
        ggtitle("Nodes in Network") +
        xlab("Time") +
        ylab("Number of Nodes") +
        norma_theme()
```


## Node Metrics

This section covers metrics that are node specific.

### Block Processing Time

The following chart shows the time required by each node for processing each individual block.

```{r block_processing_time, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
data <- filter_metric(all_data, "BlockEventAndTxsProcessingTime") %>%
    mutate(value = value / 1e6)

node_block_plot(data, "Block Processing Time", "Block Processing Time [ms]")
```

While the individual sample points may provide the highest level of detail, the smoothed out trend lines of the following plot may provide a cleaner comparison between the processing times of individual nodes.

```{r block_processing_time_smooth, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
node_block_plot(
    data, "Block Processing Time", "Block Processing Time [ms]",
    smooth = TRUE
)
```


### Block Height per Node

The following chart shows the time at which nodes have reached a given block height.

```{r block_height_per_node, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "NodeBlockStatus") %>%
    node_time_plot("Block Height over Time", "Block Height")
```

### Epoch per Node

The following chart shows the epoch number of each node over time.

```{r epoch_per_node, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "NodeEpochStatus") %>%
    node_time_plot("Epoch over Time", "Epoch")
```

### Validator Stakes

The following chart shows the received stake per validator over time.

```{r validator_stakes, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
data <- filter_metric(all_data, "ValidatorStake") %>%
    mutate(value = value / 1e18)

if (nrow(data) > 0) {
    node_time_plot(
        data, "Validator Stakes", "Stake [S]",
        xlab = "Time [s]", color_label = "Validator"
    )
}
```

### TxPool Per Node

The following charts show statistics of transaction pool.

* The number of received transactions - accumulated count over time
```{r tx_pool_per_node_received_txs, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "txpool_received") %>%
    node_time_plot("TxPool Received Transactions over Time", "Received Transactions")
```

* The number of valid transactions - accumulated count over time
```{r tx_pool_per_node_valid_txs, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "txpool_valid") %>%
    node_time_plot("TxPool Valid Transactions over Time", "Valid Transactions")
```

* The number of invalid transactions - accumulated count over time
```{r tx_pool_per_node_invalid_txs, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "txpool_invalid") %>%
    node_time_plot("TxPool Invalid Transactions over Time", "Invalid Transactions")
```

* The number of underpriced transactions - accumulated count over time
```{r tx_pool_per_node_underpriced_txs, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "txpool_underpriced") %>%
    node_time_plot(
        "TxPool Underpriced Transactions over Time", "Underpriced Transactions"
    )
```

* The number of overflowed transactions - accumulated count over time
```{r tx_pool_per_node_overflowed_txs, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "txpool_overflowed") %>%
    node_time_plot(
        "TxPool Overflowed Transactions over Time", "Overflowed Transactions"
    )
```

#### TxPool - Queued to Pending Transitions

* The number of transactions in the queued state - current count over time

```{r tx_pool_per_node_quedued_txs, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "txpool_queued") %>%
    node_time_plot("TxPool Queued Transactions over Time", "Queued Transactions")
```

* The number of transactions in the pending state - current count over time

```{r tx_pool_per_node_pending_txs, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "txpool_pending") %>%
    node_time_plot("TxPool Pending Transactions over Time", "Pending Transactions")
```

### Consumed Disk Space

Following charts show the disk space consumed by each Node. The charts
provide space consumption exclusively by the State DB only 
and the total size of the Node data directory, which contains
the State DB and other files. 

```{r state_db_size, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "statedb_disksize") %>%
    mutate(value = value / 1024 / 1024) %>%
    node_time_plot("Actual Size of StateDB over Time", "MB")
```

```{r total_db_size, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "db_size") %>%
    mutate(value = value / 1024 / 1024) %>%
    node_time_plot("Actual Size of Data Dir including StateDB over Time", "MB")
```

### System Properties

Following charts provide memory and CPU load of each Node. 

* maximal utilization of one CPU core is 100%, i.e. the CPU Load chart 
may grow above 100% depending on the number of cores. 

```{r system_cpu_procload, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "system_cpu_procload") %>%
    node_time_plot("CPU Load", "%")
```

```{r system_memory_used, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "system_memory_used") %>%
    mutate(value = value / 1024 / 1024) %>%
    node_time_plot("Memory Load", "MB")
```

## Application Metrics

### Transactions Send vs. Received

The following chart shows the total number of sent and received transactions.

```{r send_and_received_transactions_per_app, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
# Per-account incremental send counts, reused by send_rate_per_account below
sent_rate <- all_data %>%
    dplyr::filter(metric == "SentTransactions") %>%
    select(elapsed_sec, app, workers, value) %>%
    compute_incremental_rate(c("app", "workers"))

# Cumulative sent total per application
sent <- sent_rate %>%
    group_by(app) %>%
    arrange(app, elapsed_sec) %>%
    reframe(elapsed_sec = elapsed_sec, total = cumsum(rate))

# Received (committed) transactions per application
received <- filter_metric(all_data, "ReceivedTransactions")

ggplot() +
    geom_line(
        data = sent,
        aes(x = elapsed_sec, y = total, group = app,
            colour = factor(paste(app, " sent")))
    ) +
    geom_line(
        data = received,
        aes(x = elapsed_sec, y = value, group = app,
            colour = factor(paste(app, " received")))
    ) +
    ggtitle("Sent and Received Transactions per App") +
    xlab("Time") +
    ylab("Total Number of Transactions") +
    labs(colour = "Apps") +
    norma_theme()
```


The following chart shows the cummulative number of transactions send per account  (legend omitted for clarity).

```{r sent_transactions_per_app, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "SentTransactions") %>%
    mutate(account = paste(app, "_", workers)) %>%
    ggplot(aes(x = elapsed_sec, y = value,
               group = account, colour = factor(account))) +
        geom_line(show.legend = FALSE) +
        ggtitle("Sent Transactions per User") +
        xlab("Time") +
        ylab("Total Number of Transactions") +
        labs(colour = "User") +
        norma_theme()
```

The following chart shows the send-rate of individual accounts (legend omitted for clarity).

```{r send_rate_per_account, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
sent_rate %>%
    mutate(account = paste(app, "_", workers)) %>%
    ggplot(aes(x = elapsed_sec, y = rate,
               group = account, colour = factor(account))) +
        geom_line(show.legend = FALSE) +
        ggtitle("Sent Rate of Transactions per User") +
        xlab("Time") +
        ylab("Rate [Tx/s]") +
        labs(colour = "User") +
        norma_theme()
```

## Event Metrics

Metrics collected form the node emitter component.

### Emitted Events per Node

Shows the number of events emitted by each node over time. This metric can be used 
to identify whether nodes are actively participating in the network, and the effectiveness
of event throttling mechanisms.

- Solid lines represent the number of emitted events.
- Dashed lines represent the number of events skipped due to throttling.

```{r emitter_events_per_node, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
emit_data <- filter_metric(all_data, "emitter_events")
throttling_data <- filter_metric(all_data, "emitter_skip_emission_throttle")

ggplot(emit_data, aes(x = elapsed_sec, y = value, group = node, color = factor(node))) +
    geom_line(linetype = "solid") +
    geom_line(data = throttling_data, linetype = "dashed") +
    ggtitle("Emitted events over time") +
    xlab("Time") +
    ylab("Events") +
    labs(colour = "Nodes") +
    norma_theme()
```

### Transactions in Emitted Events per Node

Shows the number of transactions contained in emitted events by each node over time.
This metric can be used to identify whether nodes are effectively including transactions
in their emitted events.

```{r emitter_tx_per_node, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "emitter_txs") %>%
    node_time_plot("Transactions in emitted events over time", "Transactions")
```

### Gas used by Emitted Events per Node

Shows the amount of gas used by transactions contained in emitted events by each node over time.

```{r emitter_gas_per_node, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "emitter_gas") %>%
    node_time_plot("Gas in emitted events over time", "Gas Used")
```
