---
title: "`r if (nzchar(params$scenario)) paste0('Norma Evaluation Summary - ', params$scenario) else 'Norma Evaluation Summary'`"
output: html_document
date: "`r Sys.Date()`"
params:
    datafile: "measurements.csv"
    scenario: ""
    description: ""
    scenario_file: ""
---

```{r setup, echo=FALSE, include=FALSE}
library(tidyverse)
library(lubridate)
library(slider)

figure_dimensions <- c(10, 4)

# Load monitoring data
all_data <- read_csv(params$datafile)
start_time <- min(all_data$time, na.rm = TRUE)
if (nrow(all_data) > 0) {
    all_data <- all_data %>%
        mutate(elapsed_sec = as.numeric(time - start_time) / 1e9)
} else {
    # Guard against empty data sets to prevent crashes
    all_data$elapsed_sec <- 0
}

scenario_step_data <- all_data %>%
    dplyr::filter(metric == "ScenarioStepExecutionDuration") %>%
    transmute(
        event_name = as.character(app),
        start_sec = elapsed_sec,
        duration_sec = as.numeric(value) / 1e9
    ) %>%
    drop_na(start_sec, duration_sec) %>%
    dplyr::filter(duration_sec >= 0) %>%
    arrange(start_sec) %>%
    mutate(
        end_sec = start_sec + duration_sec,
        event_label = if_else(
            event_name == "" | is.na(event_name),
            "(unnamed event)",
            event_name
        ),
        event_idx = row_number(),
        event_fill = event_label
    )

scenario_step_fill_levels <- character(0)
scenario_step_fill_colors <- c()
scenario_step_line_colors <- c()

if (nrow(scenario_step_data) > 0) {
    event_span <- max(scenario_step_data$end_sec, na.rm = TRUE) -
        min(scenario_step_data$start_sec, na.rm = TRUE)
    min_band_width_sec <- max(0.05, event_span * 0.002)
    scenario_step_fill_levels <- unique(scenario_step_data$event_fill)
    scenario_step_fill_colors <- setNames(
        hcl.colors(length(scenario_step_fill_levels), palette = "Set 2"),
        scenario_step_fill_levels
    )
    darken_hex <- function(hex, factor = 0.55) {
        vapply(hex, function(h) {
            rgb <- grDevices::col2rgb(h)
            darker <- pmax(0, round(as.numeric(rgb) * factor))
            grDevices::rgb(
                darker[1], darker[2], darker[3],
                maxColorValue = 255
            )
        }, character(1))
    }
    scenario_step_line_colors <- setNames(
        darken_hex(scenario_step_fill_colors),
        names(scenario_step_fill_colors)
    )
    scenario_step_data <- scenario_step_data %>%
        mutate(event_fill = factor(event_fill, levels = scenario_step_fill_levels)) %>%
        mutate(event_line_color = unname(scenario_step_line_colors[as.character(event_fill)])) %>%
        mutate(visible_end_sec = pmax(end_sec, start_sec + min_band_width_sec))
} else {
    scenario_step_data$visible_end_sec <- numeric()
    scenario_step_data$event_line_color <- character()
}

# --- Helper functions ---

# Filter a named metric and convert its value column to numeric.
filter_metric <- function(data, metric_name) {
    data %>%
        dplyr::filter(metric == metric_name) %>%
        mutate(value = as.numeric(value))
}

# ggplot theme with a horizontally centered title.
norma_theme <- function() {
    theme(
        plot.title = element_text(hjust = 0.5),
        plot.background = element_blank(),
        panel.background = element_blank(),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5)
    )
}

# Scatter plot of a block metric with a smooth trend line.
# Expects columns: block, value.
block_trend_plot <- function(data, title, ylab) {
    ggplot(data, aes(x = block, y = value)) +
        geom_point() +
        geom_smooth(se = FALSE) +
        ggtitle(title) +
        xlab("Block Height") +
        ylab(ylab) +
        norma_theme()
}

# Line plot of a per-node metric over elapsed time.
# Expects columns: elapsed_sec, value, node.
node_time_plot <- function(data, title, ylab,
                           xlab = "Time",
                           color_label = "Nodes") {
    ggplot(data, aes(
        x = elapsed_sec, y = value,
        group = node, colour = factor(node)
    )) +
        geom_line() +
        ggtitle(title) +
        xlab(xlab) +
        ylab(ylab) +
        labs(colour = color_label) +
        norma_theme()
}

# Overlay faded scenario-step bands in the background for time-based charts.
# The overlay intentionally hides its own legend to avoid visual clutter.
add_scenario_step_overlay <- function(event_data = scenario_step_data) {
    if (nrow(event_data) == 0) {
        return(NULL)
    }

    list(
        geom_rect(
            data = event_data,
            aes(
                xmin = start_sec,
                xmax = visible_end_sec,
                ymin = -Inf,
                ymax = Inf,
                fill = event_fill
            ),
            inherit.aes = FALSE,
            alpha = 0.08,
            show.legend = FALSE
        ),
        geom_vline(
            data = event_data,
            aes(xintercept = start_sec),
            color = event_data$event_line_color,
            linetype = "dotted",
            linewidth = 0.3,
            show.legend = FALSE
        ),
        scale_fill_manual(values = scenario_step_fill_colors, drop = FALSE)
    )
}

# Plot of a per-node metric over block height.
# Expects columns: block, value, node.
# When smooth = TRUE, draws trend lines instead of scatter points.
node_block_plot <- function(data, title, ylab, smooth = FALSE) {
    p <- ggplot(data, aes(
        x = block, y = value,
        group = node, colour = factor(node)
    )) +
        ggtitle(title) +
        xlab("Block Height") +
        ylab(ylab) +
        labs(colour = "Nodes") +
        norma_theme()

    if (smooth) p + geom_smooth(se = FALSE) else p + geom_point()
}

# Compute a 5-second sliding window transaction rate from accumulated counts.
# data:       rows for a single metric, with numeric `time` and `value` columns
# group_cols: character vector of column names to group by
# Returns a data frame with columns: rate, time, elapsed_sec
compute_sliding_rate <- function(data, group_cols, start_time) {
    result <- data %>%
        mutate(value = as.numeric(value), time = as.numeric(time)) %>%
        group_by(across(all_of(group_cols))) %>%
        arrange(time) %>%
        reframe(
            time = time,
            increment = value - lag(value, default = 0)
        ) %>%
        arrange(time) %>%
        slide_period_dfr(
            as.POSIXct(.$time / 1e9, origin = "1970-01-01"),
            "second",
            .every = 5,
            function(d) {
                summarize(d,
                    rate = sum(increment) / 5,
                    time = (max(time) + min(time)) / 2
                )
            }
        )

    if (nrow(result) > 0) {
        result %>%
            drop_na(rate) %>%
            mutate(elapsed_sec = (time - as.numeric(start_time)) / 1e9)
    } else {
        result %>% mutate(rate = 0, elapsed_sec = 0)
    }
}

# Compute per-observation incremental rate (delta between consecutive values).
# data:       rows for a single metric with elapsed_sec and value columns
# group_cols: character vector of column names to group by
# Returns a data frame with: elapsed_sec, rate, and the grouping columns
compute_incremental_rate <- function(data, group_cols) {
    data %>%
        mutate(value = as.numeric(value)) %>%
        group_by(across(all_of(group_cols))) %>%
        arrange(across(c(all_of(group_cols), "elapsed_sec"))) %>%
        reframe(
            elapsed_sec = elapsed_sec,
            rate = value - lag(value, default = 0)
        )
}

# Compute a 5-second sliding window rate per application from accumulated counts.
# compute_sliding_rate sums the increments of everything it is given, so it is
# applied to one application at a time.
# data:             rows for a single metric with numeric `time` and `value`
#                   columns and an `app` column
# extra_group_cols: further columns the counts accumulate per, such as the
#                   individual users of an application
# Returns a data frame with columns: rate, time, elapsed_sec, app
compute_sliding_rate_per_app <- function(data, extra_group_cols, start_time) {
    apps <- sort(unique(data$app[!is.na(data$app)]))
    if (length(apps) == 0) {
        return(tibble(rate = numeric(0), elapsed_sec = numeric(0), app = character(0)))
    }

    purrr::map_dfr(apps, function(current) {
        data %>%
            dplyr::filter(app == current) %>%
            compute_sliding_rate(extra_group_cols, start_time) %>%
            mutate(app = current)
    })
}
```

This report summarizes the monitoring results of an execution of a Norma scenario.

```{r scenario_file_path, echo=FALSE, results='asis'}
if (nzchar(trimws(params$scenario_file))) {
    cat("Scenario file: ")
    cat(params$scenario_file)
    cat("\n\n")
}
```

```{r scenario_description, echo=FALSE, results='asis'}
if (nzchar(trimws(params$description))) {
    cat("## Scenario Description\n\n")
    cat(params$description)
    cat("\n")
}
```

## Scenario Execution Timeline

The following chart represents scenario execution as colored vertical bands. Each
band corresponds to one triggered scenario step; the width of the band is the
step execution duration on the driver wall clock.

```{r scenario_execution_timeline, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
if (nrow(scenario_step_data) > 0) {
    ggplot(scenario_step_data) +
        geom_rect(aes(
            xmin = start_sec,
            xmax = visible_end_sec,
            ymin = 0,
            ymax = 1,
            fill = event_fill
        ), alpha = 0.85) +
        geom_vline(
            aes(xintercept = start_sec),
            color = scenario_step_data$event_line_color,
            linetype = "dotted",
            linewidth = 0.3,
            show.legend = FALSE
        ) +
        ggtitle("Scenario Step Timeline") +
        xlab("Execution Time [s]") +
        ylab(NULL) +
        labs(fill = "Steps") +
        scale_fill_manual(values = scenario_step_fill_colors, drop = FALSE) +
        scale_y_continuous(breaks = NULL) +
        theme(
            axis.text.y = element_blank(),
            axis.ticks.y = element_blank()
        ) +
        norma_theme()
}
```

## Block Metrics

This section covers per-Block metrics which are consistent throughout the network.

### Gas per Block
The following chart shows the Gas spent per block and a trend line showing long-term trends.


```{r gas_spent_per_block, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "BlockGasUsed") %>%
    block_trend_plot("Gas per Block", "Gas spent")
```


### Transactions per Block
The following chart shows the number of Transactions per block and a trend line showing long-term trends.

```{r transactions_per_block, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "BlockNumberOfTransactions") %>%
    block_trend_plot("Transactions per Block", "Number of Transactions")
```


### Block Delay
The following chart shows the time between the earliest reported completion time of block `n-1` and earliest reported completion time block `n` for block `n`.

```{r block_delay_time, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
all_data %>%
    dplyr::filter(metric == "BlockCompletionTime") %>%
    select(block, value) %>%
    mutate(value = as.numeric(value)) %>%
    group_by(block) %>%
    summarise(value = min(value)) %>%
    arrange(block) %>%
    mutate(delta = value - lag(value)) %>%
    drop_na(delta) %>%
    ggplot(aes(x = block, y = delta / 1e9)) +
        geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
        geom_point() +
        geom_smooth(se = FALSE) +
        ggtitle("Block Delay") +
        xlab("Block Height") +
        ylab("Delay [s]") +
        norma_theme()
```


### Gas Rate per Block
The following chart shows the Gas Rate per block and a trend line.

```{r gas_rate_per_block, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "BlockGasRate") %>%
    block_trend_plot("Gas Rate per Block", "Gas per Second")
```


### Gas Base-Fee per Block
The following chart shows the Gas Base-Fee per block and a trend line.

```{r gas_base_fee_per_block, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "BlockGasBaseFee") %>%
    block_trend_plot("Gas Base Fee per Block", "Gas Base Fee")
```

## Network Metrics

This section covers metrics that are network wide properties.

### Network Incoming and Committed Transaction Rates

The following chart shows the total rate of incoming and committed transactions smoothed over 5 second intervals.

```{r throughput_over_time, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
sent_rate <- all_data %>%
    dplyr::filter(metric == "SentTransactions") %>%
    select(time, app, workers, value) %>%
    compute_sliding_rate(c("app", "workers"), start_time)

received_rate <- all_data %>%
    dplyr::filter(metric == "ReceivedTransactions") %>%
    select(time, app, value) %>%
    compute_sliding_rate("app", start_time)

ggplot() +
    geom_line(
        data = sent_rate,
        aes(x = elapsed_sec, y = rate, color = "incomming [5s window]")
    ) +
    geom_line(
        data = received_rate,
        aes(x = elapsed_sec, y = rate, color = "committed [5s window]")
    ) +
    ggtitle("Incomming and Committed Transaction Rates") +
    xlab("Time (s)") +
    ylab("Tx/s") +
    norma_theme() +
    add_scenario_step_overlay()
```


### Number of Nodes in the Network

The number of nodes in the network over time.

```{r nodes_in_network, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "NumberOfNodes") %>%
    ggplot(aes(x = elapsed_sec, y = value)) +
        geom_point() +
        ggtitle("Nodes in Network") +
        xlab("Time") +
        ylab("Number of Nodes") +
    norma_theme() +
    add_scenario_step_overlay()
```


## Node Metrics

This section covers metrics that are node specific.

### Block Processing Time

The following chart shows the time required by each node for processing each individual block.

```{r block_processing_time, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
data <- filter_metric(all_data, "BlockEventAndTxsProcessingTime") %>%
    mutate(value = value / 1e6)

node_block_plot(data, "Block Processing Time", "Block Processing Time [ms]")
```

While the individual sample points may provide the highest level of detail, the smoothed out trend lines of the following plot may provide a cleaner comparison between the processing times of individual nodes.

```{r block_processing_time_smooth, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
node_block_plot(
    data, "Block Processing Time", "Block Processing Time [ms]",
    smooth = TRUE
)
```


### Block Height per Node

The following chart shows the time at which nodes have reached a given block height.

```{r block_height_per_node, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "NodeBlockStatus") %>%
    node_time_plot("Block Height over Time", "Block Height") +
    add_scenario_step_overlay()
```

### Epoch per Node

The following chart shows the epoch number of each node over time.

```{r epoch_per_node, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "NodeEpochStatus") %>%
    node_time_plot("Epoch over Time", "Epoch") +
    add_scenario_step_overlay()
```

### Validator Stakes

The following chart shows the received stake per validator over time.

```{r validator_stakes, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
data <- filter_metric(all_data, "ValidatorStake") %>%
    mutate(value = value / 1e18)

if (nrow(data) > 0) {
    node_time_plot(
        data, "Validator Stakes", "Stake [S]",
        xlab = "Time [s]", color_label = "Validator"
    ) +
        add_scenario_step_overlay()
}
```

### TxPool Per Node

The following charts show statistics of transaction pool.

* The number of received transactions - accumulated count over time
```{r tx_pool_per_node_received_txs, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "txpool_received") %>%
    node_time_plot("TxPool Received Transactions over Time", "Received Transactions") +
    add_scenario_step_overlay()
```

* The number of valid transactions - accumulated count over time
```{r tx_pool_per_node_valid_txs, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "txpool_valid") %>%
    node_time_plot("TxPool Valid Transactions over Time", "Valid Transactions") +
    add_scenario_step_overlay()
```

* The number of invalid transactions - accumulated count over time
```{r tx_pool_per_node_invalid_txs, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "txpool_invalid") %>%
    node_time_plot("TxPool Invalid Transactions over Time", "Invalid Transactions") +
    add_scenario_step_overlay()
```

* The number of underpriced transactions - accumulated count over time
```{r tx_pool_per_node_underpriced_txs, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "txpool_underpriced") %>%
    node_time_plot(
        "TxPool Underpriced Transactions over Time", "Underpriced Transactions"
    ) +
    add_scenario_step_overlay()
```

* The number of overflowed transactions - accumulated count over time
```{r tx_pool_per_node_overflowed_txs, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "txpool_overflowed") %>%
    node_time_plot(
        "TxPool Overflowed Transactions over Time", "Overflowed Transactions"
    ) +
    add_scenario_step_overlay()
```

#### TxPool - Queued to Pending Transitions

* The number of transactions in the queued state - current count over time

```{r tx_pool_per_node_quedued_txs, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "txpool_queued") %>%
    node_time_plot("TxPool Queued Transactions over Time", "Queued Transactions") +
    add_scenario_step_overlay()
```

* The number of transactions in the pending state - current count over time

```{r tx_pool_per_node_pending_txs, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "txpool_pending") %>%
    node_time_plot("TxPool Pending Transactions over Time", "Pending Transactions") +
    add_scenario_step_overlay()
```

### Consumed Disk Space

Following charts show the disk space consumed by each Node. The charts
provide space consumption exclusively by the State DB only 
and the total size of the Node data directory, which contains
the State DB and other files. 

```{r state_db_size, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "statedb_disksize") %>%
    mutate(value = value / 1024 / 1024) %>%
    node_time_plot("Actual Size of StateDB over Time", "MB") +
    add_scenario_step_overlay()
```

```{r total_db_size, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "db_size") %>%
    mutate(value = value / 1024 / 1024) %>%
    node_time_plot("Actual Size of Data Dir including StateDB over Time", "MB") +
    add_scenario_step_overlay()
```

### System Properties

Following charts provide memory and CPU load of each Node. 

* maximal utilization of one CPU core is 100%, i.e. the CPU Load chart 
may grow above 100% depending on the number of cores. 

```{r system_cpu_procload, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "system_cpu_procload") %>%
    node_time_plot("CPU Load", "%") +
    add_scenario_step_overlay()
```

```{r system_memory_used, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "system_memory_used") %>%
    mutate(value = value / 1024 / 1024) %>%
    node_time_plot("Memory Load", "MB") +
    add_scenario_step_overlay()
```

## Transaction Lifecycle

Transactions are followed individually from the moment they are submitted to a
node until they are part of a block. They pass three phases:

- **submitted**: handed to the RPC interface of a node, which either accepted it
  into its transaction pool or refused it (*rejected*).
- **emitted**: put into an event of the DAG by a validator. Until then a
  transaction waits in a pool, either as the next nonce of its sender and
  therefore ready to be emitted (*pending*), or behind an earlier nonce of the
  same sender that is not committed yet (*stalled*).
- **included**: part of a block, and therefore executed.

### Time per Phase

One measurement per transaction is far too much to draw; the distributions are
summarised instead. The boxes span the quartiles, the whiskers reach 1.5 times
the interquartile range and everything beyond them is drawn individually. Note
the logarithmic scale: phases of a congested network differ by orders of
magnitude, not by percentages.

```{r transaction_phase_durations, echo=FALSE, message=FALSE, warning=FALSE, fig.dim = figure_dimensions}
phase_levels <- c(
    "submitted to emitted",
    "emitted to included",
    "submitted to included"
)

phase_durations <- bind_rows(
    filter_metric(all_data, "TransactionTimeToEmit") %>%
        mutate(phase = phase_levels[1]),
    filter_metric(all_data, "TransactionTimeEmitToInclude") %>%
        mutate(phase = phase_levels[2]),
    filter_metric(all_data, "TransactionTimeToInclude") %>%
        mutate(phase = phase_levels[3])
) %>%
    mutate(seconds = value / 1e9, phase = factor(phase, levels = phase_levels)) %>%
    dplyr::filter(seconds > 0)

if (nrow(phase_durations) > 0) {
    ggplot(phase_durations, aes(x = app, y = seconds, fill = phase)) +
        geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.15) +
        scale_y_log10() +
        ggtitle("Time Transactions Spent in Each Phase") +
        xlab("Application") +
        ylab("Duration [s, log scale]") +
        labs(fill = "Phase") +
        norma_theme()
} else {
    cat("No transaction durations were measured.")
}
```

```{r transaction_phase_table, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
if (nrow(phase_durations) > 0) {
    phase_summary <- phase_durations %>%
        group_by(app, phase) %>%
        summarise(
            transactions = n(),
            median = median(seconds),
            p90 = quantile(seconds, 0.9),
            p99 = quantile(seconds, 0.99),
            max = max(seconds),
            .groups = "drop"
        ) %>%
        arrange(app, phase)

    # The application whose transactions reached a block fastest; its row is the
    # one to compare the others against, so it is marked.
    total_phase <- phase_levels[3]
    best_app <- phase_summary %>%
        dplyr::filter(phase == total_phase) %>%
        slice_min(median, n = 1, with_ties = FALSE) %>%
        pull(app)

    # The table is written out directly rather than through knitr::kable, which
    # cannot rule a line between the applications nor fill a single row.
    seconds <- function(x) formatC(x, format = "f", digits = 3, big.mark = ",")
    columns <- c(
        "Application", "Phase", "Transactions",
        "Median [s]", "p90 [s]", "p99 [s]", "Max [s]"
    )
    alignments <- c("left", "left", rep("right", 5))

    header <- paste0(
        "<th style=\"padding: 3px 10px; text-align: ", alignments,
        "; border-bottom: 1px solid #333;\">", columns, "</th>",
        collapse = ""
    )

    body <- ""
    for (i in seq_len(nrow(phase_summary))) {
        row <- phase_summary[i, ]
        separator <- if (i > 1 && row$app != phase_summary$app[i - 1]) {
            "border-top: 2px solid #333;"
        } else {
            ""
        }
        highlight <- if (row$phase == total_phase && row$app == best_app) {
            "background-color: #e8f5e9;"
        } else {
            ""
        }
        values <- c(
            row$app, as.character(row$phase), format(row$transactions, big.mark = ","),
            seconds(row$median), seconds(row$p90), seconds(row$p99), seconds(row$max)
        )
        body <- paste0(
            body, "<tr>",
            paste0(
                "<td style=\"padding: 3px 10px; text-align: ", alignments, ";",
                separator, highlight, "\">", values, "</td>",
                collapse = ""
            ),
            "</tr>"
        )
    }

    cat(paste0(
        "<table style=\"border-collapse: collapse; margin-bottom: 1em;\">",
        "<caption style=\"caption-side: bottom; text-align: left; padding-top: 6px;\">",
        "Measured durations in seconds, the fastest application from submission",
        " to inclusion highlighted. The number of transactions is the number of",
        " retained measurements, which is thinned out for long runs and is",
        " therefore not the number of transactions sent.",
        "</caption>",
        "<tr>", header, "</tr>", body, "</table>"
    ))
}
```

### Time to Inclusion over Time

The same measurement over the course of the run, aggregated per interval: this
is where a growing backlog, or the moment a load stops being served, becomes
visible.

```{r transaction_latency_over_time, echo=FALSE, message=FALSE, warning=FALSE, fig.dim = figure_dimensions}
bucket_seconds <- 15

latency_over_time <- filter_metric(all_data, "TransactionTimeToInclude") %>%
    mutate(
        seconds = value / 1e9,
        bucket = floor(elapsed_sec / bucket_seconds) * bucket_seconds
    ) %>%
    dplyr::filter(seconds > 0)

if (nrow(latency_over_time) > 0) {
    ggplot(latency_over_time, aes(x = bucket, y = seconds, group = bucket)) +
        geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.15) +
        facet_wrap(~app, ncol = 1, scales = "free_y") +
        ggtitle(paste0(
            "Time from Submission to Inclusion (", bucket_seconds, "s intervals)"
        )) +
        xlab("Time") +
        ylab("Duration [s]") +
        norma_theme()
} else {
    cat("No transaction inclusion times were measured.")
}
```

### Transactions in the System

How many transactions of each application are on their way at any moment, split
by what they are waiting for. Stalled transactions are held up by their own
sender's earlier nonces, so unlike pending ones they cannot be emitted no matter
how much capacity the network has.

```{r transactions_in_system, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
state_levels <- c(
    "pending (ready to be emitted)",
    "stalled (behind an own earlier nonce)",
    "emitted (waiting for a block)"
)

in_system <- bind_rows(
    filter_metric(all_data, "TransactionsPending") %>% mutate(state = state_levels[1]),
    filter_metric(all_data, "TransactionsStalled") %>% mutate(state = state_levels[2]),
    filter_metric(all_data, "TransactionsEmitted") %>% mutate(state = state_levels[3])
) %>%
    mutate(state = factor(state, levels = state_levels))

if (nrow(in_system) > 0) {
    ggplot(in_system, aes(x = elapsed_sec, y = value, colour = state)) +
        geom_line() +
        facet_wrap(~app, ncol = 1, scales = "free_y") +
        ggtitle("Transactions in the System per Application") +
        xlab("Time") +
        ylab("Number of Transactions") +
        labs(colour = "Waiting for") +
        # No scenario-step overlay: its bands carry one colour per step, which
        # does not survive being replicated into every facet.
        norma_theme()
} else {
    cat("No transactions were tracked.")
}
```

### Committed and Rejected Transactions

Transactions that reached a block, against the ones a node refused to accept.
A rejected transaction never entered a pool, so it is lost load rather than
delayed load.

```{r transactions_committed_and_rejected, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
outcomes <- bind_rows(
    filter_metric(all_data, "TransactionsIncluded") %>% mutate(outcome = "included in a block"),
    filter_metric(all_data, "TransactionsRejected") %>% mutate(outcome = "rejected on submission")
)

if (nrow(outcomes) > 0) {
    ggplot(outcomes, aes(x = elapsed_sec, y = value, colour = app, linetype = outcome)) +
        geom_line() +
        ggtitle("Transaction Outcomes per Application") +
        xlab("Time") +
        ylab("Total Number of Transactions") +
        labs(colour = "Application", linetype = "Outcome") +
        norma_theme() +
        add_scenario_step_overlay()
} else {
    cat("No transaction outcomes were recorded.")
}
```

The same as rates, which is where a load that stops being served, or one that
starts being refused, shows up as it happens rather than as a bend in a total.

```{r transactions_committed_and_rejected_rate, echo=FALSE, message=FALSE, warning=FALSE, fig.dim = figure_dimensions}
outcome_rates <- bind_rows(
    filter_metric(all_data, "TransactionsIncluded") %>%
        select(time, app, value) %>%
        compute_sliding_rate_per_app(character(0), start_time) %>%
        mutate(outcome = "included in a block"),
    filter_metric(all_data, "TransactionsRejected") %>%
        select(time, app, value) %>%
        compute_sliding_rate_per_app(character(0), start_time) %>%
        mutate(outcome = "rejected on submission")
)

if (nrow(outcome_rates) > 0) {
    ggplot(outcome_rates, aes(x = elapsed_sec, y = rate, colour = app, linetype = outcome)) +
        geom_line() +
        ggtitle("Transaction Outcome Rates per Application [5s window]") +
        xlab("Time") +
        ylab("Tx/s") +
        labs(colour = "Application", linetype = "Outcome") +
        norma_theme() +
        add_scenario_step_overlay()
} else {
    cat("No transaction outcomes were recorded.")
}
```

### Composition of Blocks

Which application the transactions of each block came from. The bands add up to
the number of transactions of a block, so the first chart shows how much of the
network's capacity each load actually got, and the second how they shared it.
`(other)` are the transactions of no load generator - the ones Norma sends
itself to fund accounts, deploy contracts or change the rules of the network.

```{r block_composition, echo=FALSE, message=FALSE, warning=FALSE, fig.dim = figure_dimensions}
# The composition is keyed by block, while a block-to-time mapping comes from
# the completion times the nodes reported; the earliest one wins, as elsewhere.
block_times <- filter_metric(all_data, "BlockCompletionTime") %>%
    group_by(block) %>%
    summarise(
        block_sec = (min(value) - as.numeric(start_time)) / 1e9,
        .groups = "drop"
    )

composition <- filter_metric(all_data, "BlockTransactionsPerApp") %>%
    select(block, app, value) %>%
    # A block that carried nothing of an application is a zero, not a gap, or
    # the bands would be interpolated across it.
    tidyr::complete(block, app, fill = list(value = 0)) %>%
    left_join(block_times, by = "block") %>%
    drop_na(block_sec) %>%
    arrange(block)

if (nrow(composition) > 0) {
    ggplot(composition, aes(x = block_sec, y = value, fill = app)) +
        geom_area(position = "stack") +
        ggtitle("Transactions per Block by Application") +
        xlab("Time") +
        ylab("Transactions in Block") +
        labs(fill = "Application") +
        norma_theme()
} else {
    cat("No block compositions were recorded.")
}
```

```{r block_composition_share, echo=FALSE, message=FALSE, warning=FALSE, fig.dim = figure_dimensions}
if (nrow(composition) > 0) {
    ggplot(composition, aes(x = block_sec, y = value, fill = app)) +
        geom_area(position = "fill") +
        scale_y_continuous(labels = scales::percent) +
        ggtitle("Share of Each Block by Application") +
        xlab("Time") +
        ylab("Share of Block") +
        labs(fill = "Application") +
        norma_theme()
}
```

The same by gas rather than by transaction count, which is what the blocks are
actually filled with: a load of few but expensive transactions occupies a block
that looks almost empty when counting transactions.

```{r block_gas_composition, echo=FALSE, message=FALSE, warning=FALSE, fig.dim = figure_dimensions}
gas_composition <- filter_metric(all_data, "BlockGasPerApp") %>%
    select(block, app, value) %>%
    tidyr::complete(block, app, fill = list(value = 0)) %>%
    left_join(block_times, by = "block") %>%
    drop_na(block_sec) %>%
    arrange(block)

if (nrow(gas_composition) > 0) {
    ggplot(gas_composition, aes(x = block_sec, y = value / 1e6, fill = app)) +
        geom_area(position = "stack") +
        ggtitle("Gas per Block by Application") +
        xlab("Time") +
        ylab("Gas Used in Block [M]") +
        labs(fill = "Application") +
        norma_theme()
} else {
    cat("No block gas compositions were recorded.")
}
```

The same gas against the limit of the block it went into, which is the ceiling
the loads were competing for. The limit is the network's `MaxBlockGas` rule, a
technical cap a network is not normally expected to reach: what a block really
carries is governed by the gas the validators may spend per second and by the gas
one event may hold. The distance between the top of the bands and the dashed line
therefore shows how far the blocks were from that cap - usually orders of
magnitude, which is why the axis is logarithmic. A band is the gas of one
application, as above, but read on a log axis: the bands still stack up to the gas
of the block, though equal thickness no longer means equal gas.

```{r block_gas_composition_limit, echo=FALSE, message=FALSE, warning=FALSE, fig.dim = figure_dimensions}
gas_limits <- filter_metric(all_data, "BlockGasLimit") %>%
    select(block, value) %>%
    group_by(block) %>%
    summarise(gas_limit = max(value), .groups = "drop") %>%
    left_join(block_times, by = "block") %>%
    drop_na(block_sec) %>%
    arrange(block)

# A log axis has no zero, so the bands are floored below the gas of the smallest
# possible transaction; an application that contributed nothing to a block ends
# up with a band of no height, which is what it should be.
gas_floor <- 1e4
gas_stack <- gas_composition %>%
    # The transactions of no application appear in a few blocks only. Stacking
    # them last keeps them from moving the base of the stack in and out of those
    # blocks, which the log axis would exaggerate into the largest feature of the
    # chart.
    mutate(stack_last = as.integer(app == "(other)")) %>%
    arrange(block, stack_last, app) %>%
    group_by(block) %>%
    mutate(upper = cumsum(value), lower = upper - value) %>%
    ungroup()

if (nrow(gas_stack) > 0 && nrow(gas_limits) > 0) {
    ggplot(gas_stack, aes(x = block_sec)) +
        geom_ribbon(aes(
            ymin = pmax(lower, gas_floor) / 1e6,
            ymax = pmax(upper, gas_floor) / 1e6,
            fill = app
        )) +
        geom_line(
            data = gas_limits,
            aes(y = gas_limit / 1e6, linetype = "Block gas limit"),
            colour = "black",
            linewidth = 0.4
        ) +
        scale_y_log10(labels = scales::label_comma()) +
        scale_linetype_manual(values = c("Block gas limit" = "dashed")) +
        ggtitle("Gas per Block by Application against the Block Gas Limit") +
        xlab("Time") +
        ylab("Gas [M, log scale]") +
        labs(fill = "Application", linetype = NULL) +
        norma_theme()
} else {
    cat("No block gas limits were recorded.")
}
```

## Application Metrics

### Transactions Send vs. Received

The following chart shows the total number of transactions the load generators
handed to the network, and the effect they had on the applications' contracts.
Note that *received* is read from the state of those contracts rather than from
the transactions themselves - an application counting its own calls - so it
lags a submission by the time the call needs to be committed, and it counts
neither transactions that reverted nor those still on their way. The transaction
lifecycle section above measures the transactions themselves.

```{r send_and_received_transactions_per_app, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
# Per-account incremental send counts, reused by send_rate_per_account below
sent_rate <- all_data %>%
    dplyr::filter(metric == "SentTransactions") %>%
    select(elapsed_sec, app, workers, value) %>%
    compute_incremental_rate(c("app", "workers"))

# Cumulative sent total per application
sent <- sent_rate %>%
    group_by(app) %>%
    arrange(app, elapsed_sec) %>%
    reframe(elapsed_sec = elapsed_sec, total = cumsum(rate))

# Received (committed) transactions per application
received <- filter_metric(all_data, "ReceivedTransactions")

ggplot() +
    geom_line(
        data = sent,
        aes(x = elapsed_sec, y = total, group = app,
            colour = factor(paste(app, "submitted")))
    ) +
    geom_line(
        data = received,
        aes(x = elapsed_sec, y = value, group = app,
            colour = factor(paste(app, "effects on chain")))
    ) +
    ggtitle("Submitted Transactions and Their Effects per App") +
    xlab("Time") +
    ylab("Total Number of Transactions") +
    labs(colour = "Apps") +
    norma_theme() +
    add_scenario_step_overlay()
```

The same as rates: how fast an application is handing transactions over
against how fast its contract is reacting to them. The two run at the same rate
while the network keeps up, and the effects fall behind once it does not.

```{r send_and_received_rates_per_app, echo=FALSE, message=FALSE, warning=FALSE, fig.dim = figure_dimensions}
submitted_rate_per_app <- all_data %>%
    dplyr::filter(metric == "SentTransactions") %>%
    select(time, app, workers, value) %>%
    # The counts accumulate per user, and the users of one application add up to it.
    compute_sliding_rate_per_app("workers", start_time) %>%
    mutate(series = "submitted")

effect_rate_per_app <- all_data %>%
    dplyr::filter(metric == "ReceivedTransactions") %>%
    select(time, app, value) %>%
    compute_sliding_rate_per_app(character(0), start_time) %>%
    mutate(series = "effects on chain")

app_rates <- bind_rows(submitted_rate_per_app, effect_rate_per_app)

if (nrow(app_rates) > 0) {
    ggplot(app_rates, aes(x = elapsed_sec, y = rate, colour = app, linetype = series)) +
        geom_line() +
        ggtitle("Submitted Transaction and Effect Rates per App [5s window]") +
        xlab("Time") +
        ylab("Tx/s") +
        labs(colour = "Apps", linetype = "Series") +
        norma_theme() +
        add_scenario_step_overlay()
} else {
    cat("No transactions were submitted.")
}
```

The following chart shows the cummulative number of transactions send per account  (legend omitted for clarity).

```{r sent_transactions_per_app, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "SentTransactions") %>%
    mutate(account = paste(app, "_", workers)) %>%
    ggplot(aes(x = elapsed_sec, y = value,
               group = account, colour = factor(account))) +
        geom_line(show.legend = FALSE) +
        ggtitle("Sent Transactions per User") +
        xlab("Time") +
        ylab("Total Number of Transactions") +
        labs(colour = "User") +
    norma_theme() +
    add_scenario_step_overlay()
```

The following chart shows the send-rate of individual accounts (legend omitted for clarity).

```{r send_rate_per_account, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
sent_rate %>%
    mutate(account = paste(app, "_", workers)) %>%
    ggplot(aes(x = elapsed_sec, y = rate,
               group = account, colour = factor(account))) +
        geom_line(show.legend = FALSE) +
        ggtitle("Sent Rate of Transactions per User") +
        xlab("Time") +
        ylab("Rate [Tx/s]") +
        labs(colour = "User") +
    norma_theme() +
    add_scenario_step_overlay()
```

## Event Metrics

Metrics collected form the node emitter component.

### Emitted Events per Node

Shows the number of events emitted by each node over time. This metric can be used 
to identify whether nodes are actively participating in the network, and the effectiveness
of event throttling mechanisms.

- Solid lines represent the number of emitted events.
- Dashed lines represent the number of events skipped due to throttling.

```{r emitter_events_per_node, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
emit_data <- filter_metric(all_data, "emitter_events")
throttling_data <- filter_metric(all_data, "emitter_skip_emission_throttle")

ggplot(emit_data, aes(x = elapsed_sec, y = value, group = node, color = factor(node))) +
    geom_line(linetype = "solid") +
    geom_line(data = throttling_data, linetype = "dashed") +
    ggtitle("Emitted events over time") +
    xlab("Time") +
    ylab("Events") +
    labs(colour = "Nodes") +
    norma_theme() +
    add_scenario_step_overlay()
```

### Transactions in Emitted Events per Node

Shows the number of transactions contained in emitted events by each node over time.
This metric can be used to identify whether nodes are effectively including transactions
in their emitted events.

```{r emitter_tx_per_node, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "emitter_txs") %>%
    node_time_plot("Transactions in emitted events over time", "Transactions") +
    add_scenario_step_overlay()
```

### Gas used by Emitted Events per Node

Shows the amount of gas used by transactions contained in emitted events by each node over time.

```{r emitter_gas_per_node, echo=FALSE, message=FALSE, fig.dim = figure_dimensions}
filter_metric(all_data, "emitter_gas") %>%
    node_time_plot("Gas in emitted events over time", "Gas Used") +
    add_scenario_step_overlay()
```
